---
title: "Description of the NHANES dataset"
author: "Luck B. Simelus"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE)

# Load the required packages
library(pacman)
pacman::p_load(
  rio, here, tidyverse, knitr, kableExtra,
  skimr, RColorBrewer, viridis, DT, flexdashboard,
  formatR, gridExtra, janitor, ggplot2, ggpubr, cowplot, 
  gghighlight, ggokabeito
)

# Import the clean NHANES dataset
nhanes_data <- import(here("data", "cleaned_NHANES.csv"))

# Recode wave_id
nhanes_data <- nhanes_data %>%
  mutate(
    wave_id = recode(
      wave_id,
      "2015" = "2013-2014",
      "2016" = "2015-2016",
      "2017" = "2017-2018"
    )
  )
```

## Introduction

The National Health and Nutrition Examination Survey (NHANES) is a program conducted by the National Center for Health Statistics (NCHS) to assess the health and nutritional status of the U.S. population[@nhanes]. NHANES combines interviews, physical examinations, and laboratory measurements to provide information on a wide range of health indicators.

This report provides a descriptive overview of the NHANES sample across the survey waves included in this analysis. The sample is described according to its size, gender, ethnicity, and average systolic blood pressure (SBP).

## NHANES Sample composition by gender and ethnicity

```{r table-1, echo=FALSE}
# Summary table of the dataset by gender
sample_gender_summary <- nhanes_data %>%
  group_by(wave_id) %>%
  summarise(
    sample_size = n(),
    male_n = sum(gender == "Male", na.rm = TRUE),
    male_percent = mean(gender == "Male", na.rm = TRUE) * 100,
    female_n = sum(gender == "Female", na.rm = TRUE),
    female_percent = mean(gender == "Female", na.rm = TRUE) * 100
  )

# Formating with KableExtra
sample_gender_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 1,
    col.names = c(
      "NHANES wave",
      "Sample size",
      "Males (n)",
      "Males (%)",
      "Femaleales (n)",
      "Females (%)"
    ),
    caption = "Sample size and gender distribution by NHANES wave"
  ) %>%
  kable_styling(
    latex_options = "hold_position"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(1, background = "#e6f7ff", bold = TRUE)
```

The number of participants was almost similar across the three NHANES waves included in the analysis, with the 2015 wave having the most participants. Table 1 summarizes the sample size and gender distribution for each wave. Overall, males represented approximately 49% of the sample, with relatively small differences between waves.

```{r table-2, echo=FALSE}
# prepare ethnicity stats
ethnicity_summary <- nhanes_data %>%
  group_by(wave_id, ethnicity_2) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(wave_id) %>%
  mutate(
    percent = n / sum(n) * 100
  )
# sort ethnicity by overall samble size across waves
ethnicity_order <- nhanes_data %>%
  count(ethnicity_2, sort = TRUE) %>%
  pull(ethnicity_2)
# create table
ethnicity_table <- ethnicity_summary %>%
  mutate(
    ethnicity_2 = factor(ethnicity_2, levels = ethnicity_order),
    value = paste0(n, " (", round(percent, 1), "%)")
  ) %>%
  select(wave_id, ethnicity_2, value) %>%
  pivot_wider(
    names_from = wave_id,
    values_from = value
  ) %>%
  arrange(ethnicity_2)
# formating with Kableextra
ethnicity_table %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Ethnicity distribution by NHANES wave",
    col.names = c(
      "Ethnicity",
      "2013–2014",
      "2015–2016",
      "2017–2018"
    )
  ) %>%
  kable_styling(
    latex_options = "hold_position"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(1, background = "#e6f7ff", bold = TRUE)
```

The ethnic composition of the sample is presented in Table 2. The distribution of ethnic groups between survey waves was about the same.
White participants represented the largest ethnic group across the waves, followed by Black. 

## Systolic Blood Pressure
```{r table-3, echo=FALSE}
# Create average Systolic blood pressure
nhanes_data <- nhanes_data %>%
  mutate(
    avg_sbp = rowMeans(
      select(., systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4),
      na.rm = TRUE
    ),
    avg_sbp = ifelse(is.nan(avg_sbp), NA, avg_sbp)
  )
# Summary table of average sbp by wave
sbp_summary <- nhanes_data %>%
  group_by(wave_id) %>%
  summarise(
    Minimum = min(avg_sbp, na.rm = TRUE),
    Maximum = max(avg_sbp, na.rm = TRUE),
    Median = median(avg_sbp, na.rm = TRUE),
    Mean = mean(avg_sbp, na.rm = TRUE),
    SD = sd(avg_sbp, na.rm = TRUE)
  )
# Formating with Kableextra
sbp_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 1,
    caption = "Summary statistics for average systolic blood pressure by NHANES wave",
    col.names = c(
      "NHANES Wave",
      "Minimum",
      "Maximum",
      "Median",
      "Mean",
      "Standard deviation"
    )
  ) %>%
  kable_styling(
    latex_options = "hold_position"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(5, background = "#e6f7ff")
```

Table 3 summarizes the distribution of average systolic blood pressure across the NHANES waves. The mean average SBP slightly increased with each wave. The median values followed a similar pattern.

## Conclusion

Overall, the NHANES samples were relatively similar in size across the three waves, gender and ethnic composition was also relatively similar. The mean average systolic blood pressure varied somewhat between waves with the 2013-2014 wave having the lowest mean average SBP. 
These descriptive results provide an overview of the NHANES dataset and the main characteristics of some important variables for the analysis.

## References